---
title: "PCA analysis"
author: Marc Galland
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    number_sections: true
    toc_depth: 2
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
# global options
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.width = 8, 
                      fig.height = 5, 
                      fig.align = "center")
source("mypca_function.R")
suppressPackageStartupMessages(library("tidyverse"))
suppressPackageStartupMessages(library("knitr"))
```

# Data import

Two datasets:     
1. 793 metabolites' peak area detected by LC-MS from 23 different plants from 4 different genotypes.    
2. Sample to genotype correspondence.   


\newpage

## Peaks
```{r peaks}
peaks <- read.csv("../genotype_and_peak_data.csv", 
                  stringsAsFactors = F)
# The first five rows and five columns of the `peaks` dataframe. 
knitr::kable(peaks[1:5,1:5])
```


## Sample to genotype
```{r sample info}
sample_info <- read.csv("../sample_genotype_phenotype.csv", 
                        stringsAsFactors = F)
knitr::kable(sample_info)
```

\newpage

# PCA analysis (all samples)

The PCA analysis is computed with the `mypca` function that returns:
- sample scores
- variable loadings
- percentage of explained variance by each principal component (PC)

```{r pca results}
pca_results <- mypca(peaks, center = TRUE, scale = TRUE)
```

## Scree plot: variance explained
```{r scree plot}
df_explained_variance <- data.frame(
  exp_var = pca_results$explained_var$exp_var
  ) %>% 
  rownames_to_column("PC") %>% 
  mutate(PC = factor(PC,levels = unique(PC)))

scree_plot <- 
  ggplot(df_explained_variance, aes(x = PC, y = exp_var)) + 
  ylab('Explained variance (%)') + 
  ggtitle('Explained variance per principal component') + 
  geom_bar(stat = "identity")
scree_plot
```

The variance explained by PC1 is around 25%.    
PC2 and PC3 explain almost exactly the same variance (around 13%).

```{r cumulative variance}
df_explained_variance %>% 
  mutate(cumulated_variance = cumsum(exp_var)) %>% 
  ggplot(mapping = aes(x = PC, y = cumulated_variance)) +
  geom_point() +
  geom_line(group = 1) +
  labs(y = "Percentage of explained variance") +
  geom_hline(yintercept = 50, color = "red")
```

The 3 first PC explain roughly 50% of the total variance.  

\newpage

## Samples score plot
Can we distinguish the resistant genotypes from the sensitive ones on the first 3 PCs?

### PC1 versus PC2
```{r PC1 vs PC2, echo=FALSE}
scores <- pca_results$scores
scores["genotype"] <- sample_info$genotype
scores["phenotype"] <- sample_info$phenotype
scores["sample"] <- sample_info$sample

pc1_vs_pc2 <- ggplot(scores) + 
  geom_point(aes(x = PC1, y = PC2, shape = phenotype, col = genotype), size = 3) + 
  xlab(paste0('PC1 (',df_explained_variance[1,2],'% explained variance)')) + 
  ylab(paste0('PC2 (',df_explained_variance[2,2],'% explained variance)')) + 
  ggtitle('PCA score plot: PC1 versus PC2')
pc1_vs_pc2
```

Obviously, S. chm LA1840 can easily be separated on PC1 from the rest of the genotypes.  
The other genotypes are all clustered on PC1 suggesting that PC1 is not related to sensitivity or resistance perhaps. 

__Identifying the outlier on PC2__
```{r outlier PC2}
pc1_vs_pc2 +
  ggrepel::geom_text_repel(aes(x = PC1, y = PC2, label = sample)) +
  ggtitle("IL1927 sample outlier ")
```

Is it possible to separate the resistant from the sensitive genotypes?  
Let's explore the different PCs.

### PC2 versus PC3
```{r PC2 vs PC3, echo=FALSE}
pc2_vs_pc3 <- ggplot(scores) + 
  geom_point(aes(x = PC2, y = PC3, shape = phenotype, col = genotype), size = 3) + 
  xlab(paste0('PC2 (',df_explained_variance[2,2],'% explained variance)')) + 
  ylab(paste0('PC3 (',df_explained_variance[3,2],'% explained variance)')) + 
  ggtitle('PCA score plot: PC2 versus PC3')
pc2_vs_pc3
```

### PC3 versus PC4
```{r PC3 vs PC4, echo=FALSE}
pc3_vs_pc4 <- ggplot(scores) + 
  geom_point(aes(x = PC3, y = PC4, shape = phenotype, col = genotype), size = 3) + 
  xlab(paste0('PC3 (',df_explained_variance[2,2],'% explained variance)')) + 
  ylab(paste0('PC4 (',df_explained_variance[3,2],'% explained variance)')) + 
  ggtitle('PCA score plot: PC3 versus PC4')
pc3_vs_pc4
```

Not super clear but perhaps PC3 slightly separates sensitive from resistant. 

### PC3 versus PC5
```{r PC3 vs PC5, echo=FALSE}
pc3_vs_pc5 <- ggplot(scores) + 
  geom_point(aes(x = PC3, y = PC5, shape = phenotype, col = genotype), size = 3) + 
  xlab(paste0('PC3 (',df_explained_variance[2,2],'% explained variance)')) + 
  ylab(paste0('PC5 (',df_explained_variance[5,2],'% explained variance)')) + 
  ggtitle('PCA score plot: PC3 versus PC5')
pc3_vs_pc5
```

### PC4 versus PC5
```{r PC4 vs PC5, echo=FALSE}
pc4_vs_pc5 <- ggplot(scores) + 
  geom_point(aes(x = PC4, y = PC5, shape = phenotype, col = genotype), size = 3) + 
  xlab(paste0('PC4 (',df_explained_variance[4,2],'% explained variance)')) + 
  ylab(paste0('PC5 (',df_explained_variance[5,2],'% explained variance)')) + 
  ggtitle('PCA score plot: PC4 versus PC5')
pc4_vs_pc5
```


Still not very clear and PC5 explains only 6% of the total variance so will stop here. 

\newpage

# PCA analysis (IL27_6 removed)

## Outlier removal

From the previous analysis, two individuals appeared different from the rest of their 
corresponding genotype:
- IL27_6  
- s_ch_1  

Since IL27_6 appears very different from the other IL1927 samples, it will be removed before
the PCA analysis is done. Same for s_ch_1.

```{r outlier removal, echo=FALSE}
# finding the row index of IL27_6
row_index_of_outliers = c(
  match(x = "IL27_6", table = sample_info$sample),
  match(x = "s_ch_1", table = sample_info$sample)
  )

# remove this line from the peaks data
peaks_wo_outliers = peaks[-row_index_of_outliers,]

# remove it from the sample to genotype df
sample_info_wo_outliers <- sample_info[-row_index_of_outliers,]

sample_info_wo_outliers <- sample_info_wo_outliers %>%
  mutate(genotype = factor(genotype, levels = c("KG1955", "IL1928", "IL1927", "LA1840")))
  
knitr::kable(sample_info_wo_outliers)
```

The original `peaks` dataframe has 23 rows and the filtered `peaks_wo_outlier` dataframe has now 21 rows. 
Same for the samples

## Scree plot (wo outlier)
```{r pca wo outlier}
pca_results2 <- mypca(peaks_wo_outliers, center = TRUE, scale = TRUE)

df_explained_variance2 <- data.frame(
  exp_var = pca_results2$explained_var$exp_var
  ) %>% 
  rownames_to_column("PC") %>% 
  mutate(PC = factor(PC,levels = unique(PC)))

scree_plot2 <- 
  ggplot(df_explained_variance2, aes(x = PC, y = exp_var)) + 
  ylab('Explained variance (%)') + 
  ggtitle('Explained variance per principal component') + 
  geom_bar(stat = "identity")
scree_plot2
ggsave("scree_plot.png")
```

## Samples score plot (wo outlier)


### PC1 versus PC2
```{r PC1 vs PC2 wo outlier, echo=FALSE}
palette <- c("#E69F00", "#D55E00", "#0072B2", "#56B4E9")

scores2 <- pca_results2$scores
scores2["genotype"] <- sample_info_wo_outliers$genotype
scores2["phenotype"] <- sample_info_wo_outliers$phenotype
scores2["sample"] <- sample_info_wo_outliers$sample

pc1_vs_pc2 <- ggplot(scores2) + 
  geom_point(aes(x = PC1, y = PC2, shape = phenotype, col = genotype), size = 4) + 
  xlab(paste0('PC1 (',df_explained_variance2[1,2],'% explained variance)')) + 
  ylab(paste0('PC2 (',df_explained_variance2[2,2],'% explained variance)')) + 
  scale_color_manual(values = palette) +
  scale_fill_manual(values = palette) +
  theme_simple()
pc1_vs_pc2
ggsave("sample_score_plot.png")
```

This plot seems to show a much clearer picture with S. chm LA1840 clearly separable from 
the other genotypes on PC1 (as before).

But now, it seems that IL1927 (resistant) is also more intermediate between LA1840 (resistant) and the other sensitive genotypes (KG1955 "elite line", IL1927).   

\newpage

## Loadings (wo outliers)

Since PC1 is now more related to our phenotype of interest (resistance/sensitivity), we can extract the metabolites with the highest loadings on PC1. 

```{r loadings, echo=TRUE}
loadings <- pca_results2$loadings

loadings_long <- loadings %>% 
  rownames_to_column("metabolite") %>% 
  select(metabolite, PC1) %>% 
  mutate(abs_PC1 = abs(PC1)) %>% 
  arrange(desc(abs_PC1))
kable(head(loadings_long, n = 10))
```

We can also visualise it as a barplot. 

```{r}
loadings_long %>% 
  arrange(desc(abs_PC1)) %>% 
  top_n(10) %>% 
  ggplot(., aes(x = metabolite, y = PC1)) +
  geom_bar(stat = "identity") 
```

