---
title: "data_exploration"
author: "Lissy Denkers"
date: "1/4/2022"
output: html_document
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.width = 8, 
                      fig.height = 5, 
                      fig.align = "center")
suppressPackageStartupMessages(library("ggplot2"))
suppressPackageStartupMessages(library("ggrepel"))
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("skimr"))
suppressPackageStartupMessages(library("Hmisc"))
suppressPackageStartupMessages(library("tidyr"))
suppressPackageStartupMessages(library("car"))
suppressPackageStartupMessages(library("rstatix"))
suppressPackageStartupMessages(library("tidyverse"))
suppressPackageStartupMessages(library("broom"))
suppressPackageStartupMessages(library("drc"))
suppressPackageStartupMessages(library("modelr"))
suppressPackageStartupMessages(library("egg"))
```

# Data import 

## Background information

This is the data of the whitefly development bioassay on grafts of MM and LA1840. The nymphs in all developmental stages (first to fourth instar) were counted every other day. After the fourth instar stage, whiteflies develop into adults, leaving behind the larval skin called exuviae. Nymphs in the last phase of fourth instar stage and exuviae were removed from the leaf after each count to prevent a whitefly outbreak in the greanhouse.

See metadata file for more information.

## The raw data
```{r flies}
flies <- read.csv("raw_data.csv", 
                         header = T, 
                         sep = ",", 
                         dec = ".", 
                         check.names = FALSE)
# The first 10 rows of the data
knitr::kable(flies[1:10,1:8])
```

## Preparing the data for analysis

A few improvements to the data should be made before analysis.

First, three plants must be excluded from analysis, based on observations during counting (see metadata file).

```{r filter}
flies <- flies %>%     
  dplyr::filter(place != 4) %>%     
  dplyr::filter(place < 13)
```

Next, I make data wide for access to separate life stages.
```{r wide}
flies <- flies %>%
  mutate(stage = as.factor(stage)) %>%
  pivot_wider(names_from = stage, values_from = number) 
```

Because the fourth instars are the end point and taken of after counting, combine fourth instars and exuviae and make their numbers cumulative over time. 
```{r total fourth instars}
flies <- flies %>%
  mutate(fourth_exuviae = `5_late_fourth_instar` + `6_exuviea`) %>%
  group_by(place) %>%
  mutate(`5_completed_lifecycle` = cumsum(fourth_exuviae)) %>%
  mutate(`6_total_fourth_instars` = `4_early_fourth_instar` + `5_completed_lifecycle`) %>%
  ungroup()
```

Remove separate and non-cumulative late fourth instar and exuviae columns.
```{r remove unnessesary columns}
flies <- flies %>%
  select(-`5_late_fourth_instar`, -`6_exuviea`, -fourth_exuviae)
```

Now the data can be made long again.
```{r long}
flies <- flies %>%
  pivot_longer(cols = `0_egg`:`6_total_fourth_instars`, 
               names_to = "stage", values_to = "number")
```

Add the relative number of nymphs as percentage from the number of eggs and as percentage from the number of hatched eggs
```{r relative numbers}
flies <- flies %>%
  mutate(percentage_from_eggs = number / eggs_start * 100) %>%
  mutate(percentage_from_hatched = number / hatched * 100)

```


Add a new dataframe for the total number of nymphs per day without percentage, based on "flies"
```{r flies_wide}
flies_wide <- flies  %>%
  select(-percentage_from_eggs) %>%
  select(-percentage_from_hatched)
```

and make the data wide.
```{r wide}
flies_wide <- flies_wide %>%
  mutate(stage = as.factor(stage)) %>%
  pivot_wider(names_from = stage, values_from = number) 
```

Add the total number of nymphs per sample per day and the relative total as percentage from the number of eggs and as percentage from the number of hatched eggs
```{r}
flies_wide <- flies_wide %>%
  mutate(total = rowSums(flies_wide[,8:12], na.rm = TRUE)) %>%
  mutate(percentage_from_eggs = total / eggs_start * 100) %>%
  mutate(percentage_from_hatched = total / hatched * 100)
```

# Different developmental stages, different effects

In the original dataset of Arjen with ungrafted plants, you can see different effects of LA1840 on the different whitefly lifestages. The number of eggs was higher on LA1840 than on MM, while the number of fourth instar nymphs was drasticly decreased on LA1840. While the number of eggs is probably the result of the effect the plant has on adult flies, the nymph development is more likely to be influenced by the diet of the nymphs themselves.
Another effect the plants could have, is affecting the hatching of the eggs. This can however not be analysed from Arjens data.

Because I want to have a clear view of those different effects, I will go through each part separately.

## Eggs

On normal LA1840 plants, the number of eggs is increased compared to MM plants. Can we see the same effect when an LA1840 rootstock is used with a MM scion?

First a quick overview of the data:
```{r}
flies_wide %>%
  group_by(genotype) %>%
  get_summary_stats(`0_egg`, type = "mean_sd")

```

And now visualized in a boxplot:
```{r egg boxplot, echo=FALSE, message=FALSE, warning=FALSE}
flies_wide %>%  
  ggplot(., aes(x = genotype, y = `0_egg`)) +
  geom_boxplot()  +
  geom_jitter(size = 1, alpha = 0.5, width = 0.1) +
  labs(y="number of eggs at start", 
       x= "rootstock genotype") +
  theme_classic()
```

It seems like the number of eggs might be higher on the grafts with an LA1840 rootstock.

Before testing this, check the assumptions:
- normality
- homogeneity of variance
- no significant outliers

```{r warning = FALSE}

shapiro.test(resid(aov(flies_wide$`0_egg`~flies_wide$genotype)))
```
The p-value of the Shapiro-Wilk test is >0.05, so the data is normally distributed

```{r warning = FALSE}
leveneTest(`0_egg`~genotype, data = flies_wide)
```
The Levene's test also has p>0.05, so the variance is equal between the two groups.

Now the outliers:
```{r warning = FALSE}
flies_wide %>%
  group_by(genotype) %>%
  identify_outliers(`0_egg`)
```
There is one outlier, but it is not extreme.

Because all assumptions for a t-test are met, we can do a Students t-test.
```{r warning = FALSE}
flies_wide %>%
  t_test(`0_egg` ~ genotype, var.equal = TRUE) %>%
  add_significance()
```

```{r warning = FALSE}
flies_wide %>%
  cohens_d(`0_egg` ~ genotype, var.equal = TRUE)
```

The difference in number of eggs on MM and LA1840 grafts is significant (p=0.0468) and the effectsize of the rootstock genotype on the number of eggs is large (d=1.39).

The d-value indicates the difference between the two means in number of standard deviations. So the mean of LA1840 grafts is 1.39 standard deviations higher than that of MM grafts.

### Effect of rootstock genotype on the number of eggs

Similar to the previous findings on normal LA1840 and MM plants, the number of eggs are increased on grafts with an LA1840 rootstock compared to grafts with a MM rootstock. This indicates that the signal responsible for the increased oviposition is transported from the LA1840 rootstock to the MM scion.


## Hatching

Is there an effect of rootstock genotype on the hatching of the eggs? For this I will use the absolute number of hatched eggs, as well as the percentage.

### absolute number of hatched eggs

An overview of the data:
```{r warning = FALSE}
flies_wide %>%  
  mutate(day = as.factor(day)) %>%
  group_by(genotype) %>%
  dplyr::filter(day=="2") %>%
  get_summary_stats(hatched, type = "mean_sd")
```

```{r warning = FALSE}
flies_wide %>%  
  mutate(day = as.factor(day)) %>%
  dplyr::filter(day == "2") %>% 
  ggplot(., aes(x = genotype, y = hatched)) +
  geom_boxplot()  +
  geom_jitter(size = 1, alpha = 0.5, width = 0.1) +
  labs(y="number of hatched eggs",
       x= "rootstock genotype") +
  theme_classic()
```
Shapiro-Wilk test:
```{r warning = FALSE}
flies_wide %>%  
  mutate(day = as.factor(day)) %>%
  dplyr::filter(day == "2") %>%
  group_by(genotype) %>%
  shapiro_test(hatched)
```
Levene's test
```{r warning = FALSE}
flies_wide %>%  
  mutate(day = as.factor(day)) %>%
  dplyr::filter(day == "2") %>%
  levene_test(hatched~genotype)
```
Identifying extreme outliers:
```{r warning = FALSE}
flies_wide %>%  
  mutate(day = as.factor(day)) %>%
  dplyr::filter(day == "2") %>%
  group_by(genotype) %>%
  identify_outliers(hatched)
```

The assumptions for a t-test are met.
```{r warning = FALSE}
flies_wide %>%  
  mutate(day = as.factor(day)) %>%
  dplyr::filter(day == "2") %>%
  t_test(hatched ~ genotype, var.equal = TRUE) %>%
  add_significance()
```
```{r warning = FALSE}
flies_wide %>%  
  mutate(day = as.factor(day)) %>%
  dplyr::filter(day == "2") %>%
  cohens_d(hatched ~ genotype, var.equal = TRUE)
```

Again, there is a significant difference between LA1840 and MM grafts (p=0.04) with a large effectsize of the rootstock genotype (d=1.45).
This could, however, be caused by the difference in number of eggs. Therefore, I will also check the success of hatching relative to the amount of eggs.

### Percentage of hatching

An overview of the data:
```{r warning = FALSE}
flies_wide <- flies_wide %>%
  mutate(perc_hatched = hatched/eggs_start*100) %>%  
  mutate(day = as.factor(day))

flies_wide %>%
  group_by(genotype) %>%
  dplyr::filter(day=="2") %>%
  get_summary_stats(perc_hatched, type = "mean_sd")
```
```{r}
flies_wide %>%
  dplyr::filter(day == "2") %>% 
  ggplot(., aes(x = genotype, y = perc_hatched)) +
  geom_boxplot()  +
  geom_jitter(size = 1, alpha = 0.5, width = 0.1) +
  labs(y="hatched eggs (%)",
       x= "rootstock genotype") +
  theme_classic()
```

Shapiro-Wilk test:
```{r warning = FALSE}
flies_wide %>% 
  dplyr::filter(day == "2") %>%
  group_by(genotype) %>%
  shapiro_test(perc_hatched)
```

Levene's test
```{r warning = FALSE}
flies_wide %>% 
  dplyr::filter(day == "2") %>%
  levene_test(perc_hatched~genotype)
```

Identifying extreme outliers:
```{r warning = FALSE}
flies_wide %>%
  dplyr::filter(day == "2") %>%
  group_by(genotype) %>%
  identify_outliers(perc_hatched)
```

The assumptions for a t-test are met.
```{r warning = FALSE}
flies_wide %>%
  dplyr::filter(day == "2") %>%
  t_test(perc_hatched ~ genotype, var.equal = TRUE) %>%
  add_significance()
```

```{r warning = FALSE}
flies_wide %>% 
  dplyr::filter(day == "2") %>%
  cohens_d(perc_hatched ~ genotype, var.equal = TRUE)
```

There is no difference in the percentage of hatched eggs between the two rootstock genotypes.
There is, however, a large variation in hatching between plants, as can be seen in the above boxplot.


```{r}
flies_wide %>%
  select(!(`1_first_instar`:perc_hatched)) %>%  
  dplyr::filter(day == "2") %>%
  pivot_longer(cols = hatched:`0_egg`, 
               names_to = "stage", values_to = "number") %>%
  mutate(place = as.factor(place)) %>%
  anova_test(number ~ stage*genotype + Error(place/stage)) %>%
  get_anova_table()
```
The same result is visible in a two-way repeated measures anova. Although the number of eggs and of hatched eggs differs between the rootstock genotypes and the number of hatched eggs is lower than the initial number of eggs, there is no genotype*stage effect. This means that the rootstock genotype does not influence the hatching rate.


### Effect of rootstock genotype on hatching

Although the LA1840 grafts have a higher number of hatched eggs, the grafting itself does not seem to be influenced by the rootstock, as the percentage of hatched eggs is equal on the LA1840 and MM grafts. 

## Nymph development

For the nymph development, I want to focus on the development from first instar to fourth instar. To cancel-out effect on oviposition and variation in hatching, take percentage of later nymphs from first instars.

First, I'll make a new dataframe for the total number of nymphs that passed a developmental stage  for each plant
```{r}
# new data set from flies_wide without unnecessary columns
flies_total <- flies_wide %>%
  select(!c(`0_egg`, `1_first_instar`, `4_early_fourth_instar`, `5_completed_lifecycle`, total:perc_hatched)) 

# create new columns for the number of nymphs passing a developmental stage
flies_total <- flies_total %>%
  mutate(second_instars = rowSums(flies_total[,7:9])) %>%
  mutate(third_instars = rowSums(flies_total[,8:9])) %>%
  mutate(fourth_instars = `6_total_fourth_instars`)

# remove unnecessary columns
flies_total <- flies_total %>%
  select(!(`2_second_instar`:`6_total_fourth_instars`))

# make data long
flies_total <- flies_total %>%
  pivot_longer(cols = eggs_start:fourth_instars, 
               names_to = "stage", values_to = "number")
 
# make data wide for day
flies_total <- flies_total %>%
  select(-date) %>%
  pivot_wider(names_from = day, values_from = number)  

# make a column for the highest number per developmental stage per plant over time
flies_total <- flies_total %>%
  rowwise() %>%
  mutate(number = max(c_across(`2`:`29`)))

# remove the counts per day
flies_total <- flies_total %>%
  select(!c(`2`:`29`))

# order the developmental stages
flies_total$stage <- factor(flies_total$stage, 
                          levels= c("eggs_start", "hatched", "second_instars", 
                                    "third_instars", "fourth_instars"))

# The first 10 rows of the data
knitr::kable(flies_total[1:10,1:4])
```

And also for the numbers relative to eggs and hatched eggs
```{r}
# new  wide dataframe from flies_total for numbers relative to eggs
flies_relative_eggs <- flies_total %>%
  pivot_wider(names_from = stage, values_from = number)

# make a column for each stage relative to eggs
flies_relative_eggs <- flies_relative_eggs %>%
  mutate(eggs = eggs_start/eggs_start*100) %>%
  mutate(hatched_eggs = hatched/eggs_start*100) %>%
  mutate(second = second_instars/eggs_start*100) %>%
  mutate(third = third_instars/eggs_start*100) %>%
  mutate(fourth = fourth_instars/eggs_start*100)

# remove absolute numbers and make long
flies_relative_eggs <- flies_relative_eggs %>%
  select(!c(eggs_start:fourth_instars)) %>%
  pivot_longer(cols = eggs:fourth, 
               names_to = "stage", values_to = "perc_from_eggs")

# order the developmental stages
flies_relative_eggs$stage <- factor(flies_relative_eggs$stage, 
                          levels= c("eggs", "hatched_eggs", "second", 
                                    "third", "fourth"))

# The first 10 rows of the data
knitr::kable(flies_relative_eggs[1:10,1:4])
```

```{r}
# new  wide dataframe from flies_total for numbers relative to hatched
flies_relative_hatched <- flies_total %>%
  pivot_wider(names_from = stage, values_from = number)

# make a column for each stage relative to hatched
flies_relative_hatched <- flies_relative_hatched %>%
  mutate(eggs = eggs_start/hatched*100) %>%
  mutate(hatched_eggs = hatched/hatched*100) %>%
  mutate(second = second_instars/hatched*100) %>%
  mutate(third = third_instars/hatched*100) %>%
  mutate(fourth = fourth_instars/hatched*100)

# remove absolute numbers and make long
flies_relative_hatched <- flies_relative_hatched %>%
  select(!c(eggs_start:fourth_instars)) %>%
  pivot_longer(cols = eggs:fourth, 
               names_to = "stage", values_to = "perc_from_hatched")

# order the developmental stages
flies_relative_hatched$stage <- factor(flies_relative_hatched$stage, 
                          levels= c("eggs", "hatched_eggs", "second", 
                                    "third", "fourth"))

# The first 10 rows of the data
knitr::kable(flies_relative_hatched[1:10,1:4])
```

What happens in the absolute numbers through the development?

```{r}
flies_total %>%  
  mutate(number = na_if(number, 0)) %>%
  group_by(genotype, stage) %>% 
  ggplot(., aes(x = stage, y = number)) +
  facet_wrap( vars(genotype), nrow = 1) +
  geom_boxplot(aes(color = genotype))   +
  geom_jitter(size = 1, alpha = 0.5, width = 0.1) +
  labs(y="absolute number of nymphs that developed past stage",
       x= "developmental stage") +
  theme_classic()
```

The number of nymphs appears to be higher on LA1840 grafts, but this could be the result of the higher number of eggs, as discussed above.

What happens if we look at the number of nymphs relative to the number of hatched eggs?

```{r}
flies_relative_hatched %>%    
  dplyr::filter(stage != "eggs") %>%    
  dplyr::filter(stage != "hatched_eggs") %>%  
  group_by(genotype, stage) %>% 
  ggplot(., aes(x = stage, y = perc_from_hatched)) +
  facet_wrap( vars(genotype), nrow = 1) +
  geom_boxplot(aes(color = genotype))   +
  geom_jitter(size = 1, alpha = 0.5, width = 0.1) +
  labs(y="number of nymphs relative to hatched (%)",
       x= "developmental stage") +
  theme_classic()
```

Now it seems like a larger percentage of eggs which hatched on MM grafts developed into fourth instar nymph than on LA1840 grafts.

Is this a significant difference?

An overview of the data:
```{r warning = FALSE}
flies_relative_hatched %>%
  group_by(genotype) %>%
  dplyr::filter(stage=="fourth") %>%
  get_summary_stats(perc_from_hatched, type = "mean_sd")
```

Shapiro-Wilk test:
```{r warning = FALSE}
flies_relative_hatched %>% 
  dplyr::filter(stage=="fourth") %>%
  group_by(genotype) %>%
  shapiro_test(perc_from_hatched)
```

Levene's test
```{r warning = FALSE}
flies_relative_hatched %>%
  dplyr::filter(stage=="fourth") %>%
  levene_test(perc_from_hatched~genotype)
```

Identifying extreme outliers:
```{r warning = FALSE}
flies_relative_hatched %>%
  dplyr::filter(stage=="fourth") %>%
  group_by(genotype) %>%
  identify_outliers(perc_from_hatched)
```

The assumptions for a t-test are met.
```{r warning = FALSE}
flies_relative_hatched %>%
  dplyr::filter(stage=="fourth") %>%
  t_test(perc_from_hatched ~ genotype, var.equal = TRUE) %>%
  add_significance()
```

```{r warning = FALSE}
flies_relative_hatched %>%
  dplyr::filter(stage=="fourth") %>%
  cohens_d(perc_from_hatched ~ genotype, var.equal = TRUE)
```

The percentage of hatched eggs that develop into fourth instar nymphs is indeed lower on LA1840 grafts than on MM grafts (p=0.046). The rootstock genotype has a large effect (d=1.40) on the development of nymphs into fourth instars after hatching.


### Effect of rootstock genotype on the nymph development

In line with the previous results on normal LA1840 and MM plants, the nymph development is decreased on the LA1840 grafts compared to the MM grafts. Contrary to the previous findings, however, this is only visible in the development relative to the number of hatched eggs.


## First conclusions

overall similar to results of Arjen, but diluted.
Two contradicting effects of LA1840 rootstock on whiteflies: 
- increased oviposition
- hampered nymph development

Effects found on MM scion suggests transportable nature of responsible mechanism(s).


# Phenotype in detail

Now, let's have a more detailed look at the development.

### Hatching (first instars)


