---
title: "nymph plots"
author: "Lissy Denkers"
date: "`r Sys.Date()`"
output: 
  pdf_document: default
  html_document: default
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.width = 8, 
                      fig.height = 5, 
                      fig.align = "center")
suppressPackageStartupMessages(library("ggplot2"))
suppressPackageStartupMessages(library("ggrepel"))
suppressPackageStartupMessages(library("ggpubr"))
suppressPackageStartupMessages(library("ggpubfigs"))
suppressPackageStartupMessages(library("patchwork"))
suppressPackageStartupMessages(library("plyr"))
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("skimr"))
suppressPackageStartupMessages(library("Hmisc"))
suppressPackageStartupMessages(library("tidyr"))
suppressPackageStartupMessages(library("car"))
suppressPackageStartupMessages(library("rstatix"))
suppressPackageStartupMessages(library("tidyverse"))
suppressPackageStartupMessages(library("broom"))
suppressPackageStartupMessages(library("drc"))
suppressPackageStartupMessages(library("modelr"))
suppressPackageStartupMessages(library("egg"))
suppressPackageStartupMessages(library("tidydrc"))
suppressPackageStartupMessages(library("aomisc"))
suppressPackageStartupMessages(library("devtools"))
```

# data import

## The raw data
```{r flies}
flies <- read.csv("raw_data.csv", 
                         header = T, 
                         sep = ",", 
                         dec = ".", 
                         check.names = FALSE)
# The first 10 rows of the data
knitr::kable(flies[1:10,1:8])
```


```{r filter}
flies <- flies %>%     
  dplyr::filter(place != 4) %>%     
  dplyr::filter(place < 13)
```

```{r wide}
flies <- flies %>%
  mutate(stage = as.factor(stage)) %>%
  pivot_wider(names_from = stage, values_from = number) 
```
 
```{r total fourth instars}
flies <- flies %>%
  mutate(fourth_exuviae = `5_late_fourth_instar` + `6_exuviea`) %>%
  group_by(place) %>%
  mutate(`5_completed_lifecycle` = cumsum(fourth_exuviae)) %>%
  mutate(`6_total_fourth_instars` = `4_early_fourth_instar` + `5_completed_lifecycle`) %>%
  ungroup()
```

```{r graph}
relative_flies <- flies %>%
  dplyr::select(-`4_early_fourth_instar`: -`5_completed_lifecycle`)

relative_flies <- relative_flies %>%
  mutate(first_instars = rowSums(relative_flies[,8:11])) %>%
  mutate(second_instars = rowSums(relative_flies[,9:11])) %>%
  mutate(third_instars = rowSums(relative_flies[,10:11])) %>%
  mutate(fourth_instars = `6_total_fourth_instars`)

relative_flies <- relative_flies %>%
  group_by(place) %>%
  mutate(first = `1_first_instar` / max(eggs_start))%>%
  mutate(second = `2_second_instar` / max(first_instars))%>%
  mutate(third = `3_third_instar` / max(second_instars))%>%
  mutate(fourth = fourth_instars / max(third_instars)) %>%
  ungroup()

relative_flies <- relative_flies %>%
  dplyr::select(-eggs_start: -fourth_instars)

relative_flies <- relative_flies %>%
  pivot_longer(cols = first:fourth, 
               names_to = "stage", values_to = "number") %>%
  mutate(genotype = factor(genotype, levels = c("MM", "LA1840"))) 
```

```{r test_flies}
test_flies <- relative_flies
# 10 rows of middle of the data
knitr::kable(test_flies[197:207,1:6])
```

## adjust data

```{r wide}
test_flies <- test_flies %>%
  mutate(stage = as.factor(stage)) %>%
  mutate(place = as.factor(place)) %>%
  group_by(stage, genotype) %>%
  pivot_wider(names_from = place, values_from = number) 

# 10 rows of middle of the data
knitr::kable(test_flies[41:51,1:15])
``` 

```{r median SD}
test_flies <- test_flies %>%
  mutate(day = as.factor(day)) %>%
  rowwise() %>%
  mutate(mean_number = mean(c_across(`1`:`12`), na.rm = TRUE)) %>%
  mutate(median_number = median(c_across(`1`:`12`), na.rm = TRUE)) %>%
  mutate(sd_number = sd(c_across(`1`:`12`), na.rm = TRUE)) %>%
  mutate(max_number = max(c_across(`1`:`12`), na.rm = TRUE)) %>%
  mutate(min_number = min(c_across(`1`:`12`), na.rm = TRUE))

test_flies <- test_flies %>%
  dplyr::select(-`1`: -`12`)

# 10 rows of middle of the data
knitr::kable(test_flies[41:51,1:9])
```

# graphs development relative to previous stage

graph with observed minima en maxima (dotted lines), confidence interval (median +/- SD; filled area) and median (solid line) per genotype per day for each stage, relative to the previous stage.
```{r first instars}

pfirst <- test_flies %>%
  dplyr::filter(stage == "first") %>% 
  mutate(day = as.numeric(day)) %>% 
  group_by(genotype) %>% 
  ggplot() +
  theme_simple() +
  labs(y = "first",
       x = "days after infection") +
  scale_color_manual(values = palette) +
  scale_fill_manual(values = palette) +
  geom_line(aes(x = day,
                  y = min_number,
                  color = genotype),
              linetype = "dotted") +
  geom_line(aes(x = day,
                  y = max_number,
                  color = genotype),
              linetype = "dotted") +
  geom_ribbon(aes(x = day,
                  ymin = (median_number - sd_number),
                  ymax = (median_number + sd_number),
                  fill = genotype),
              alpha = 0.35) +
  geom_line(aes(x = day,
                y = median_number,
                color = genotype),
            size = 1.2)

```

```{r second instars}
psecond <- test_flies %>%
  dplyr::filter(stage == "second") %>% 
  mutate(day = as.numeric(day)) %>% 
  group_by(genotype) %>% 
  ggplot() +
  theme_simple() +
  labs(y = "second",
       x = "days after infection") +
  scale_color_manual(values = palette) +
  scale_fill_manual(values = palette) +
  geom_line(aes(x = day,
                  y = min_number,
                  color = genotype),
              linetype = "dotted") +
  geom_line(aes(x = day,
                  y = max_number,
                  color = genotype),
              linetype = "dotted") +
  geom_ribbon(aes(x = day,
                  ymin = (median_number - sd_number),
                  ymax = (median_number + sd_number),
                  fill = genotype),
              alpha = 0.35) +
  geom_line(aes(x = day,
                y = median_number,
                color = genotype),
            size = 1.2)

```

```{r third instars}
pthird <- test_flies %>%
  dplyr::filter(stage == "third") %>% 
  mutate(day = as.numeric(day)) %>% 
  group_by(genotype) %>% 
  ggplot() +
  theme_simple() +
  labs(y = "Third",
       x = "days after infection") +
  scale_color_manual(values = palette) +
  scale_fill_manual(values = palette) +
  geom_line(aes(x = day,
                  y = min_number,
                  color = genotype),
              linetype = "dotted") +
  geom_line(aes(x = day,
                  y = max_number,
                  color = genotype),
              linetype = "dotted") +
  geom_ribbon(aes(x = day,
                  ymin = (median_number - sd_number),
                  ymax = (median_number + sd_number),
                  fill = genotype),
              alpha = 0.35) +
  geom_line(aes(x = day,
                y = median_number,
                color = genotype),
            size = 1.2) 

```

```{r fourth instars}
pfourth <- test_flies %>%
  dplyr::filter(stage == "fourth") %>% 
  mutate(day = as.numeric(day)) %>% 
  group_by(genotype) %>% 
  ggplot() +
  theme_simple() +
  labs(y = "fourth",
       x = "days after infection") +
  scale_color_manual(values = palette) +
  scale_fill_manual(values = palette) +
  geom_line(aes(x = day,
                  y = min_number,
                  color = genotype),
              linetype = "dotted") +
  geom_line(aes(x = day,
                  y = max_number,
                  color = genotype),
              linetype = "dotted") +
  geom_ribbon(aes(x = day,
                  ymin = (median_number - sd_number),
                  ymax = (median_number + sd_number),
                  fill = genotype),
              alpha = 0.35) +
  geom_line(aes(x = day,
                y = median_number,
                color = genotype),
            size = 1.2)

(pfirst + psecond) / (pthird + pfourth)  / guide_area() +
  plot_layout(guides = 'collect')
```

Add cumulative graphs and try with relative abundance


## cummulative numbers

This is what happens for the cummulative numbers:

```{r graph}
relative_cummulative <- flies %>%
  dplyr::select(-`4_early_fourth_instar`: -`5_completed_lifecycle`)

relative_cummulative <- relative_cummulative %>%
  mutate(first_instars = rowSums(relative_cummulative[,8:11])) %>%
  mutate(second_instars = rowSums(relative_cummulative[,9:11])) %>%
  mutate(third_instars = rowSums(relative_cummulative[,10:11])) %>%
  mutate(fourth_instars = `6_total_fourth_instars`)

relative_cummulative <- relative_cummulative %>%
  group_by(place) %>%
  mutate(first = first_instars / max(eggs_start))%>%
  mutate(second = second_instars / max(first_instars))%>%
  mutate(third = third_instars / max(second_instars))%>%
  mutate(fourth = fourth_instars / max(third_instars)) %>%
  ungroup()

relative_cummulative <- relative_cummulative %>%
  dplyr::select(-eggs_start: -fourth_instars)

relative_cummulative <- relative_cummulative %>%
  pivot_longer(cols = first:fourth, 
               names_to = "stage", values_to = "number") %>%
  mutate(genotype = factor(genotype, levels = c("MM", "LA1840"))) 

p1 <- relative_cummulative %>%
  dplyr::filter(stage == "first") %>% 
  dplyr::filter(day < 15) %>% 
  group_by(genotype) %>% 
  ggplot(aes(x=day, y=number, color = genotype)) +
  labs(y = "first instars",
       x = "days after infection") +
  geom_point() +
  stat_smooth(method = drm, method.args = list(fct = W1.3()), se = FALSE) +
  scale_color_manual(values = palette) +
  theme_simple()

p2 <- relative_cummulative %>%
  dplyr::filter(stage == "second") %>% 
  dplyr::filter(day < 20) %>% 
  group_by(genotype) %>% 
  ggplot(aes(x=day, y=number, color = genotype)) +
  labs(y = "second instars",
       x = "days after infection") +
  geom_point() +
  stat_smooth(method = drm, method.args = list(fct = LL.3()), se = FALSE) +
  scale_color_manual(values = palette) +
  theme_simple()

p3 <- relative_cummulative %>%
  dplyr::filter(stage == "third") %>% 
  group_by(genotype) %>% 
  ggplot(aes(x=day, y=number, color = genotype)) +
  labs(y = "third instars",
       x = "days after infection") +
  geom_point() +
  stat_smooth(method = drm, method.args = list(fct = LL.3()), se = FALSE) +
  scale_color_manual(values = palette) +
  theme_simple()

p4 <- relative_cummulative %>%
  dplyr::filter(stage == "fourth") %>% 
  group_by(genotype) %>% 
  ggplot(aes(x=day, y=number, color = genotype)) + 
  labs(y = "fourth instars",
       x = "days after infection") +
  geom_point() +
  stat_smooth(method = drm, method.args = list(fct = LL.3()), se = FALSE) +  
  scale_color_manual(values = palette) +
  theme_simple()

(p1 + p2) / (p3 + p4)  / guide_area() +
  plot_layout(guides = 'collect')
```
