---
title: "targeted riboflavin analysis of 4 chmielewskii genotypes"
author: "Lissy Denkers"
date: "`r Sys.Date()`"
output: 
  pdf_document: default
  html_document: default
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.width = 8, 
                      fig.height = 5, 
                      fig.align = "center")
suppressPackageStartupMessages(library("ggplot2"))
suppressPackageStartupMessages(library("ggpubr"))
suppressPackageStartupMessages(library("ggpubfigs"))
suppressPackageStartupMessages(library("ggsignif"))
suppressPackageStartupMessages(library("patchwork"))
suppressPackageStartupMessages(library("multcomp"))
suppressPackageStartupMessages(library("car"))
suppressPackageStartupMessages(library("tidyverse"))
suppressPackageStartupMessages(library("rstatix"))
suppressPackageStartupMessages(library("broom"))
suppressPackageStartupMessages(library("DescTools"))
```

# Data import 

## Background information

This is the targeted analysis of riboflavin in the leaves and phloem of the plants used in the whitefly development bioassay on S. chmielewskii accesions LA1840, LA1028, LA1330 and LA2663 and S. lycopersicum cv MM. The data is normalised to an internal standard (a stable isotope of riboflavin) and displayed in nM (based on a riboflavin standard calibration curve).

## The raw data
```{r flies}
flies <- read.csv("riboflavin_leaf_phloem_cleaned.csv", 
                         header = T, 
                         sep = ",", 
                         dec = ".", 
                         check.names = FALSE) %>%  
  mutate_if(is.character, str_replace_all, 
            pattern = "MM", replacement = "cv")  %>%
  dplyr::filter(genotype != "cv") 
  

# The first 10 rows of the data
knitr::kable(flies[1:10,1:4])
```

```{r wide}
flies <- flies %>%
  mutate(tissue = as.factor(tissue)) %>%
  pivot_wider(names_from = tissue, values_from = riboflavin) 
```

```{r alter the units} 
#flies <- flies %>%
#  dplyr::mutate(leaf = leaf/1000) %>% # set as nmol/mL leaf extract
#  dplyr::mutate(phloem = phloem*2) # set as pmol/mL phloem exudate

```

```{r genotype order}
flies <- flies %>%
  mutate(genotype = factor(genotype, levels = c("LA1028", "LA1330", "LA1840", "LA2663"))) 
print(flies)
```


# plots
```{r color palette}
palette <- c("#bbe1f6", "#88caef", "#56B4E9", "#4490ba")

# "#E69F00", for cv
```

```{r check color palette}
library(colorblindcheck)
palette_check(palette, plot=TRUE)
```

These statistics are added to the plots:
```{r leaf anova}
  leaf_ribo <- aov(flies$leaf ~ flies$genotype)
  summary(leaf_ribo)
  
  EtaSq(leaf_ribo)
  
  TukeyHSD(leaf_ribo, conf.level=.95) 
```

```{r calculate median, SD and SE leaf}
flies %>%
  group_by(genotype) %>%
  dplyr::summarise(mean = mean(leaf), SD = sd(leaf), 
            SE = sd(leaf)/sqrt(n()))
```


```{r phloem anova}
  phloem_ribo <- aov(flies$phloem ~ flies$genotype)
  summary(phloem_ribo)
  
  EtaSq(phloem_ribo)
  
  TukeyHSD(phloem_ribo, conf.level=.95) 
```

```{r calculate median, SD and SE phloem}
flies %>%
  group_by(genotype) %>%
  dplyr::summarise(mean = mean(phloem), SD = sd(phloem), 
            SE = sd(phloem)/sqrt(n()))
```


```{r combined plot, fig.width = 16, fig.height = 10}
p1 <- flies %>% 
  ggplot(aes(x=genotype, y=leaf)) + 
  labs(y = "Riboflavin in leaf material (pmol/mg fresh weight)") +
  geom_boxplot(aes(fill = genotype)) +   
  geom_point(aes(fill = genotype), size = 2, shape = 21,
             position = position_jitterdodge(jitter.width = 1.4)) + 
  theme_simple() + 
  scale_fill_manual(values = palette)  +  
  scale_y_continuous(expand = c(0, 0), limits = c(0, 7.5))  +
  geom_signif(comparison = list(c("LA1330","LA1840")), 
              map_signif_level = TRUE, 
              annotations = c("."),
              y_position = 7,
              vjust = -0.5,
              textsize = 7)

p2 <- flies %>% 
  ggplot(aes(x=genotype, y=phloem)) + 
  labs(y = "Riboflavin in phloem exudates (pmol/mL)") +
  geom_boxplot(aes(fill=genotype)) +   
  geom_point(aes(fill = genotype), size = 2, shape = 21,
             position = position_jitterdodge(jitter.width = 1.4)) + 
  theme_simple() + 
  scale_fill_manual(values = palette)   +  
  scale_y_continuous(expand = c(0, 0), limits = c(0, 7.5))  +
  geom_signif(comparison = list(c("LA1840", "LA2663")), 
              map_signif_level = TRUE, 
              annotations = c("."),
              y_position = 7,
              vjust = -0.1,
              textsize = 10)

 

collected <- (p2) / guide_area() +
  plot_layout(guides = 'collect') 
#  plot_annotation(tag_levels = 'A')

collected

ggsave(filename = "4_chmielewskii_only_riboflavin.png", plot = collected, width = 5, height = 10)
ggsave(filename = "4_chmielewskii_ony_riboflavin.svg", plot = collected, width = 5, height = 10)
```

